Polynomial Chaos for the Estimation of Manning's Friction

### 2013-12-16 paper/analysis.tex
more edits for results

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
We now use the surrogate created effectively using PC to propagate prior 
uncertainty through the forward model. We mainly exploit the surrogate 
to study the statistics of the water surface elevation, and to conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.
We note that the statistical and sensitivity analysis presented below 
is performed using the PC coefficients solely; these were calculated
from an ensemble of 125 \geoclaw simulations.
\subsubsection{Statistical analysis}
The mean of a QoI and its standard deviation is computed
from the PC coefficients as shown in Equation~\ref{eq:mean} and Equation~\ref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations in Figure~\ref{fig:rlzs}.
We observed that the standard deviation ......

@@

\subsubsection{Sensitivity analysis}
A global sensitivity analysis is next conducted to quantity the contribution of each
uncertain parameter to the variance in water surface elevation. To this end, we calculate 
the total sensitivity index using the PC coefficients as shown in Equation~\eqref{eq:T-hard}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
The Manning's roughness coefficient at the shore $N_2$ is clearly dominant and contributes
the most to the variance in the water surface elevation compared to the other two 
Manning's roughness coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
and at the four gauges. The Manning's roughness coefficient
in the bottom of ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
during the second hour of simulation and Manning's roughness coefficient
at the land $N_1$ appears to be an insignificant contributor
to the variance.

In 2D, the sensitivity analysis shows also that $N_2$ is dominant
where we have variance in water surface elevation. This is
indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
for $N_1$(top row) $N_2$(center row) and $N_3$ (bottom row)
at different times as indicated in each panel.

\subsubsection{Response surface}
In addition to the static and analysis, the PC surrogate 
can be used to construct a response surface for the uncertain input parameters.
This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
range. Since, $N_3$ shows insignificant contribution in the 
uncertainty in the model output, we find the impact of $N_2$ and $N_3$ for 
fixed $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
and Figure~\ref{fig:response3}
for different times as indicated. The most striking features are the relatively flat
horizontal contours in the $N_3$ direction indicating that water surface elevation depends
only mildly on $N_3$ even during peak tsunami events.

>>>>

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
The PC expansion created using an ensemble of 125 \geoclaw simulations
is now used as a surrogate to propagate prior input parameter uncertainty 
through the forward model. We here exploit the surrogate 
to study the statistics of the water surface elevation and conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the calculated PC mean water surface elevation along with two standard deviations
bounds at the four gauges as indicated in each panel.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observations is that the
standard deviation in water surface elevations vanishes at few times instances
during the tsunami. This can be explained as ......

@@

\subsubsection{Sensitivity analysis}
A global sensitivity analysis is next performed to quantity the contribution of each
uncertain parameter to the variance in water surface elevation. To this end, we calculate 
the total sensitivity index using the PC coefficients as shown in Equation~\eqref{eq:T-hard}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
The Manning's roughness coefficient at the shore $N_2$ is clearly dominant and contributes
the most to the variance in the water surface elevation compared to the other two 
Manning's roughness coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
and at the four gauges. The Manning's roughness coefficient
in the bottom of ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
during the second hour of simulation and Manning's roughness coefficient
at the land $N_1$ appears to be an insignificant contributor
to the variance.

In 2D, the sensitivity analysis shows also that $N_2$ is dominant
at the same regions where we observe variance in water surface elevation. This is
indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
for $N_1$(top row) $N_2$(center row) and $N_3$ (bottom row)
at different times as indicated in each panel.

\subsubsection{Response surface}
In addition to the statistical and sensitivity analysis, the PC surrogate 
can be used to construct a response surface for the uncertain input parameters.
This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
range. Since $N_3$ shows insignificant contribution in the 
uncertainty in the model output, we find the impact of $N_2$ and $N_3$ for 
fixed value of $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
horizontal contours in the $N_3$ direction indicating that water surface elevation depends
only mildly on $N_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
at gauge 21419, its the opposite case where the horizontal contours are in the $N_2 $ direction
and there water surface elevation depends mildly on $N_2$.

### 2013-12-16 paper/mcmc.tex
more edits for results

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one.
To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles
with their counterparts from the observed gauge locations.
The plots show good agreement between the simulations and the observations at different
locations and at different times. Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}.

We now can exploit the surrogate model in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent \geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

>>>>

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one. To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles (with error bars)
with their counterparts from the observed gauge locations.
The plots show a good agreement between the simulations and the 
observations at different locations and at different times. 
The same comparison can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}. The difference between the observations
and simulations is attributed to the uncertainty in the 
Manning's N coefficient and to the measurement errors to be quantified
in this section.

The surrogate model is exploited in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent \geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

### 2013-12-16 paper/pcstudy.tex
more edits for results

A final check consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC representation.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3}
The different curves correspond to increased order of PC, from 1 to 5.
The plots indicate that the double peaked distributions are
well-resolved with order 4 but becomes weakly insenstive to further refinement 
as shown with order 5. The various error metrics presented above 
provide confidence that the PC expansion is a faithful 
model surrogate.

>>>>

A final check consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC representation.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3}
The different curves correspond to increased order of PC (from 1 to 5).
The plots indicate that the double peaked distributions are
well-resolved with order 3 at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
However, at t = 10800 s further refinement is needed as the pdfs are
sensitive to the refinement up to order 4 but then becomes weakly 
insensitive with additional refinement  (order 5) as indicated in Figure~\ref{fig:pdfs3}. 
We, therefore, used PC order 5 in all computations.
This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
model surrogate that can be used in both the forward and inverse problems..
